---
title: "Exploration of Highly Correlated Simulated Data"
author: "Dan Manela"
date: "24/11/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison of Methods}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>",
  cache = FALSE,
  fig.width=7, 
  fig.height=7
)
# knitr::knit_theme$set("earendel")
```

As always, we begin by loading the package.

```{r load, include=TRUE, message=FALSE}
library(causl)
library(ggplot2)
library(purrr)
library(rjson)
library(survey)
library(ivreg)
library(tidyverse)
```

## Context

In this example we work with the following model:
$$Z~\sim~\mathcal{N}(0, 1) \\ X~|~Z~\sim~\mathcal{N}( \rho_{XZ} Z,~1 - \rho_{XZ}^{2}) \\ Y~|~\text{do}(X) \sim \mathcal{N}(X - 0.5,~1)$$
with a gaussian copula correlation of $\rho_{YZ}=0.5$.

## Outcome Regression

```{r}
data <- list()
rhos <- c(0.2, 0.4, 0.6, 0.8)
i <- 1
for (path in list.files('validation_datasets/multivar_gaussian/', pattern='gaussian_cop', full.names=TRUE)) {
  print('================================================')
  print(paste0('For dataset: ', path))
  print('================================================')  
  print('')
  
  cop_data <- suppressMessages(read_csv(path))
  or_rho <- suppressMessages(lm(Y ~ X + Z, data=cop_data))

  results <- as_tibble(summary(or_rho)$coef) %>%
    mutate(true_coefs=c(-0.5, rhos[i], 0.5)) %>%
    mutate(ub=Estimate + 2 * `Std. Error`) %>%
    mutate(lb=Estimate - 2 * `Std. Error`) %>%
    mutate(
      is_true_coef_in_estimate_range=if_else(
        (ub> true_coefs) & (lb < true_coefs)
      , TRUE, FALSE)
    ) %>%
    dplyr::select(-`t value`, `Pr(>|t|)`)
  print(results)
  i <- i + 1
}
```
The true regression coefficients (estimated theoretically) lie within the empirical regression estimates and their uncertainty estimates.

## IPW

Due to the low overlap observed for larger $\rho$, IPW returns incorrect regression estimates for this exercise. However, the correct estimates via correctly specified Outcome Regression show that the generated data is indeed being drawn from our desired distribution.  

```{r}
data <- list()
paths <- list.files('validation_datasets/multivar_gaussian/', pattern='gaussian_cop', full.names=TRUE)
rhos <- c(0.2, 0.4, 0.6, 0.8)

for (i in 1:length(paths)) {
  path <- paths[i]
  rho <- rhos[i]
  print('================================================')
  print(paste0('For dataset: ', path))
  print('================================================')  
  print('')
  cop_data <- suppressMessages(read_csv(path))
  
  mod_num <- lm(X ~ 1, data=cop_data)
  cop_data <- cop_data %>% mutate(
    num=dnorm(
      x=cop_data$X, 
      mean=fitted.values(mod_num),
      sd=summary(mod_num)$sigma
    )
  )
  mod_den <- lm(X ~ Z, data=cop_data)
  cop_data <- cop_data %>% mutate(
    den=dnorm(
      x=cop_data$X,
      mean=fitted.values(mod_den),
      sd=summary(mod_den)$sigma
    )
  )
  cop_data <- cop_data %>% mutate(wt=num/den)

  glm_rho <- svyglm(
    Y ~ X, 
    design = svydesign(id=~1, data=cop_data, weights = ~cop_data$wt)
    )

  results <- as_tibble(summary(glm_rho)$coef) %>%
    mutate(true_coefs=c(-0.5, rho)) %>%
    mutate(
      is_true_coef_in_estimate_range=if_else(
        ((`Estimate` + 2 * `Std. Error`) > true_coefs) & ((`Estimate` - 2 * `Std. Error`) < true_coefs)
      , TRUE, FALSE)
    ) %>%
    dplyr::select(-`t value`, `Pr(>|t|)`)
  print(results)
}
```
